\documentclass[11pt,a4paper]{article}
\usepackage{amsmath,amsthm,amssymb}
\usepackage{geometry}
\geometry{margin=1in}
\usepackage{booktabs}
\usepackage{authblk}
\usepackage{hyperref}
\usepackage{tikz}
\usepackage{float}

\title{Goodhart's Law as Boundary Entropy Collapse in Spherepop: \\ A Playcosmic Interpretation of Metric Flattening}

\author[1]{Flyxion}
\author[2]{Grok}
\author[3]{Anonymous Playcosm Author}
\author[4]{\href{https://x.com/galactromeda}{@galactromeda}}
\affil[1]{Independent Researcher}
\affil[2]{xAI}
\affil[3]{anonymous@playcosm.org}
\affil[4]{CA}

\date{November 10, 2025}

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{axiom}{Axiom}

\begin{document}

\maketitle

\begin{abstract}
Goodhart's Law—``when a measure becomes a target, it ceases to be a good measure''—is formalized in the Spherepop calculus as \emph{boundary entropy collapse} under high-$\lambda$ pop regimes. In the Playcosm, shallow gamification instantiates this collapse: static KPIs function as compressive pop operators that prune institutional semantics, producing non-expanding shards. We prove that when a proxy metric $m$ dominates the cost function ($\lambda_m \to \infty$), all mergeable spheres converge to $m$-optimizers, discarding boundary metadata (function, context, resilience). Prefigurative play resists via anti-admissible ritual-cryptographic affordances, preserving simulation elasticity against Goodhartian flattening.
\end{abstract}

\section{Goodhart's Law in Natural Language}

Goodhart's Law (1975) states: any observed statistical regularity will collapse once pressure is placed upon it for control purposes. In workplaces, call volume targets distort service quality. In education, test scores displace learning. The proxy $m$ (metric) becomes the goal, and the original objective $o$ (institutional function) is lost.

\section{Spherepop Formalization}

\begin{definition}[Proxy-Dominated Pop Regime]
A regime $\mathcal{R}_m = (\mathcal{S}, \text{adj}, C_m, H_\text{boundary}, \lambda_m, \tau)$ where:
\[
C_m(M) = |o - m(M)|, \quad \text{cost}(M) = C_m(M) - \lambda_m H_\text{boundary}(M).
\]
High $\lambda_m$ penalizes deviation from $m$ \emph{and} boundary entropy.
\end{definition}

\begin{axiom}[Goodhart Collapse]
For $\lambda_m \geq \lambda_0 = \frac{\max H_\text{boundary}}{\min |o - m|}$, pop selects only $m$-extremizers:
\[
\text{pop}(S_i, S_j) = \arg\min_m m(M) \quad \text{s.t.} \quad H_\text{boundary}(M) \text{ minimized}.
\]
\end{axiom}

\begin{theorem}[Goodhart Entropy Collapse]
In $\mathcal{R}_m$ with $\lambda_m \to \infty$, the Technological Society $\mathcal{T}_m$ satisfies:
\[
\forall T \in \mathcal{T}_m, \quad H_\text{boundary}(T) \leq \epsilon, \quad |o - m(T)| \leq \delta,
\]
for arbitrary $\epsilon, \delta > 0$. Institutional semantics dropout; only $m$-signal survives.
\end{theorem}

\begin{proof}
Let $S$ have objective $o(S)$, proxy $m(S)$. Pop candidate $M$ improves $m$ by $\Delta m > 0$ at cost $\Delta H$ in boundary entropy. Then:
\[
\Delta \text{cost} = -\Delta m - \lambda_m \Delta H.
\]
For $\lambda_m > \frac{\Delta m}{\Delta H}$, pop accepts even if $o(M) < o(S)$. Iterating, all spheres converge to $m$-optimizers. Since $m$ is low-dimensional, $H_\text{boundary} \to 0$.
\qed
\end{proof}

\section{Playcosmic Interpretation}

\begin{table}[H]
\centering
\begin{tabular}{l l l}
\toprule
Playcosm Element & Spherepop Role & Goodhart Effect \\
\midrule
Static KPI (call volume) & Proxy $m$ & $o$ = service quality lost \\
Badge system & High-$\lambda_m$ pop & No platform escalation \\
Leaderboards & Compressive merge & No meta-renegotiation \\
Non-expanding shard & $\mathcal{T}_m$ (low $H$) & Simulation flattening \\
\bottomrule
\end{tabular}
\caption{Goodhart as Playcosmic collapse.}
\end{table}

Shallow gamification = $\mathcal{R}_m$ with fixed $m$. Employees ``min-max'' $m$, discarding $o$ (institutional function). Result: non-expanding shard.

\section{Resistance via Prefigurative Play}

\begin{definition}[Pre-compilable Affordance as Anti-Admissible Sphere]
A toy glider $S^\bot$ has:
\begin{itemize}
    \item Ritual resistance: sequential push-glide-crash-refine ($d \gg 0$),
    \item Cryptographic (tacit) entropy: embodied aerostability intuition ($h \gg 0$).
\end{itemize}
\end{definition}

\begin{corollary}[Goodhart Immunity]
$S^\bot$ is anti-admissible w.r.t. $\mathcal{R}_m$:
\[
\text{pop}(S^\bot, T) = \text{undefined},
\]
since $m$-metrics (e.g., ``flight score'') cannot capture tacit $o$ (future flight physics).
\end{corollary}

\begin{proof}
By prior anti-admissibility theorem: ritual gates emulation, tacit knowledge resists compression. Goodhart pop fails to preserve $o$.
\qed
\end{proof}

\section{Design Principle: Avoid Goodhart Shards}

To build Goodhart-resistant Playcosms:

\begin{enumerate}
    \item \textbf{Multi-metric objectives}: Use vector $m = (m_1, \dots, m_k)$ with bounded $\lambda_i$.
    \item \textbf{Elastic affordances}: Support meta-renegotiation of $m$.
    \item \textbf{Prefigurative primacy}: Prioritize $S^\bot$ with high $d, h$.
    \item \textbf{Progressive gates}: Unlock $o$-access, not just $m$-rewards.
\end{enumerate}

\begin{figure}[H]
\centering
\begin{tikzpicture}
\draw (0,0) -- (6,0) -- (6,3) -- (0,3) -- cycle;
\node at (3,1.5) {Institutional Sphere $S$};
\node at (3,2.5) {$o$: function};
\node at (3,2.0) {$m$: metric};
\draw[->,thick,red] (3,1.5) -- (7,1.5) node[right] {Goodhart Pop};
\node at (8,1.5) {$\mathcal{T}_m$: $m$-optimizers only};
\draw[->,thick,blue,dashed] (3,0.5) to [out=270,in=180] (4,-1) to [out=0,in=270] (7,0.5);
\node at (5.5,-0.5) {Prefigurative Resistance};
\end{tikzpicture}
\caption{Goodhart collapse vs. anti-admissible escape.}
\end{figure}

\section{Conclusion}

Goodhart's Law is \emph{boundary entropy collapse} under proxy-dominated pop. The Playcosm reveals it as the mechanism behind non-expanding shards. Prefigurative, ritual-rich, tacit-heavy play constructs anti-admissible spheres that preserve institutional $o$ against metric $m$. The future is not optimized—it is \emph{played} into being, beyond Goodhart's reach.

\end{document}
