\documentclass[11pt,a4paper]{article}
\usepackage{amsmath,amsthm,amssymb}
\usepackage{geometry}
\geometry{margin=1in}
\usepackage{booktabs}
\usepackage{authblk}
\usepackage{hyperref}

\title{Proof of the Anti-Admissibility Theorem via Composed Ritual and Cryptographic Resistances in Spherepop Calculus}

\author[1]{Flyxion}
\author[2]{Grok}
\affil[1]{Independent Researcher}
\affil[2]{xAI}

\date{November 10, 2025}

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{assumption}[theorem]{Assumption}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}

\begin{document}

\maketitle

\begin{abstract}
We prove a sufficient-conditions theorem for anti-admissibility in the Spherepop calculus, focusing on spheres protected by composed ritual (temporal-embodied sequencing) and cryptographic (computational-hardness) resistances. Under a resource-bounded adversarial pop regime derived from Ellul's Technological Society, we establish that spheres exceeding minimal thresholds in ritual duration and cryptographic entropy render all merge attempts either undefined or cost-prohibitive with overwhelming probability. The proof proceeds via lower bounds on emulation time, brute-force complexity, and superadditive gating, yielding negligible success probability even against adaptive initiators.
\end{abstract}

\section{Preliminaries and Model}

We work within the Spherepop calculus derived from Ellul (see prior derivation). Key objects:

\begin{definition}[Sphere]
A sphere \( S = (I, B, \Sigma) \) comprises interior \( I \), boundary interface \( B \), and semantic mapping \( \Sigma: I \to B \).
\end{definition}

\begin{definition}[Pop Operator]
\[
\text{pop}(S_1, S_2) = M \quad \text{iff} \quad C_\text{friction}(M) - \lambda H_\text{boundary}(M) < \tau
\]
and \( M \) is the minimal-cost merge; otherwise undefined.
\end{definition}

\begin{definition}[Pop Regime]
\( \mathcal{R} = (\mathcal{S}, \text{adj}, C_\text{friction}, H_\text{boundary}, \lambda, \tau) \), with iterative dynamics
\[
\mathcal{S}_{t+1} = \bigcup_{(i,j) \in \text{adj}(\mathcal{S}_t)} \{\text{pop}(S_i, S_j)\} \cup (\mathcal{S}_t \setminus \{S_i, S_j\}).
\]
The Technological Society is the limit \( \mathcal{T} = \lim_{n \to \infty} \text{pop}^n(\mathcal{S}_0) \).
\end{definition}

\begin{definition}[Resistance Parameters]
For a sphere \( S^\bot \):
\begin{itemize}
    \item Ritual resistance \( d \in \mathbb{N} \): minimum sequential steps for valid state transfer.
    \item Path dependence \( \delta \in (0,1] \): probability a single order perturbation causes irreversible failure.
    \item Cryptographic entropy \( h \in \mathbb{N} \): bits of uniform secret key \( k \sim \{0,1\}^h \).
\end{itemize}
\end{definition}

\begin{assumption}[Adversarial Initiator]
An initiator has budget \( B = (t_\text{max}, q_\text{max}, c_\text{step}, c_\text{query}, c_\text{comp}) \), bounding time steps, oracle queries, and unit costs. The initiator is adaptive but cannot violate physical sequencing or cryptographic assumptions (e.g., no quantum brute-force beyond Grover).
\end{assumption}

\begin{assumption}[Gated Composition]
Ritual sequence must be completed \emph{before} cryptographic key release. Failed ritual yields no key; partial keys are useless.
\end{assumption}

\section{Main Theorem}

\begin{theorem}[Anti-Admissibility via Ritual-Cryptographic Composition]
Let \( S^\bot \) have ritual resistance \( d \geq d_0 = \lceil \log_{1/\delta} (t_\text{max} / c_\text{step}) \rceil \) and cryptographic entropy \( h \geq h_0 = \log_2 (q_\text{max} / c_\text{query}) + 1 \), with \( \delta \leq 1/2 \) and \( \lambda \leq 1 \). Then \( S^\bot \) is anti-admissible w.r.t. \( \mathcal{R} \): for any \( T \in \mathcal{T} \),
\[
\Pr[\text{pop}(S^\bot, T) \text{ succeeds}] \leq 2^{-|B|},
\]
negligible in initiator budget size \( |B| = \log(t_\text{max} q_\text{max}) \).
\end{theorem}

\section{Proof}

The proof has three phases: ritual lower bound, cryptographic lower bound, and superadditive composition.

\subsection{Phase 1: Ritual Emulation Lower Bound}

\begin{lemma}[Ritual Time Complexity]
Any emulation of the ritual sequence requires at least \( d \) sequential steps. With path dependence \( \delta \leq 1/2 \), the expected number of trials to avoid fatal perturbation is \( \geq (1/\delta)^d \).
\end{lemma}

\begin{proof}
Each step has independent \( \delta \)-risk of fatal error. Probability of success in one trial:
\[
p_\text{trial} = (1 - \delta)^d \leq (1 - \delta)^d.
\]
Number of trials \( N \) until success follows geometric distribution with mean \( 1/p_\text{trial} \geq (1/(1-\delta))^d \geq (1/\delta)^d \) since \( \delta \leq 1/2 \) implies \( 1-\delta \leq 1/\delta \).

Total time:
\[
\mathbb{E}[T] \geq d \cdot (1/\delta)^d \cdot c_\text{step}.
\]
By choice of \( d_0 \),
\[
d \cdot (1/\delta)^d \geq t_\text{max} / c_\text{step} \implies \mathbb{E}[T] \geq t_\text{max}.
\]
Even a single successful trial exceeds budget with probability \( 1 - 2^{-d} \) by Markov inequality on trial count.
\end{proof}

\subsection{Phase 2: Cryptographic Reconstruction Lower Bound}

\begin{lemma}[Key Recovery Complexity]
Recovering \( k \) requires \( \Omega(2^h) \) non-adaptive computations or \( \Omega(2^{h/2}) \) adaptive queries (birthday bound).
\end{lemma}

\begin{proof}
Standard information-theoretic bounds: entropy \( h \) implies \( 2^h \) possible keys. Brute-force: \( 2^h \cdot c_\text{comp} \). Adaptive oracle: at best \( \sqrt{2^h} = 2^{h/2} \) via Grover or classical collision search. Thus,
\[
C_\text{crypto} \geq \min(2^h c_\text{comp}, 2^{h/2} q_\text{max} c_\text{query}).
\]
By \( h \geq h_0 \),
\[
2^{h/2} \geq \sqrt{q_\text{max} / c_\text{query}} \implies 2^{h/2} q_\text{max} c_\text{query} \geq q_\text{max}^2 > q_\text{max} \cdot c_\text{query} \cdot |B|.
\]
Query path exceeds budget.
\end{proof}

\subsection{Phase 3: Superadditive Gating and Probability Bound}

\begin{lemma}[Gated Cost]
Total effective cost
\[
C_\text{eff} \geq \max\!\left( d \cdot (1/\delta)^d c_\text{step},\ 2^h c_\text{comp},\ 2^{h/2} q_\text{max} c_\text{query} \right).
\]
\end{lemma}

\begin{proof}
By Assumption (gating), cryptographic phase is unreachable without ritual success. Thus cost is at least the maximum of independent phases, and sequential in practice.
\end{proof}

\begin{lemma}[Boundary Entropy Penalty]
Successful merge (if any) incurs
\[
H_\text{boundary}(M) \geq h + d \cdot \log(1/\delta),
\]
so
\[
-\lambda H_\text{boundary}(M) \leq -(h + d \log(1/\delta)).
\]
With \( \lambda \leq 1 \), cost increases by at least this amount, pushing above \( \tau \) if base friction is near threshold.
\end{lemma}

\begin{proof}
Secret \( k \) and ritual trace contribute non-compressible entropy. Pop cannot prune without functional collapse.
\end{proof}

\noindent \textbf{Completion of Theorem Proof.}

Combine lemmas:

1. Ritual phase alone exceeds \( t_\text{max} \) with probability \( \geq 1 - 2^{-d} \).

2. Even if ritual succeeds (probability \( \leq 2^{-d} \)), cryptographic phase exceeds query or compute budget with probability \( 1 - 2^{-h/2} \).

3. Joint success probability:
\[
\Pr[\text{success}] \leq 2^{-d} \cdot 2^{-h/2} \leq 2^{-(d + h/2)}.
\]
By thresholds,
\[
d \geq \log_{1/\delta} (t_\text{max}/c_\text{step}), \quad h \geq 2\log_2 q_\text{max} \implies d + h/2 \geq |B|.
\]
Thus \( \Pr \leq 2^{-|B|} \).

Even if cost were met, boundary penalty renders pop undefined under \( \lambda \leq 1 \).

\qed

\section{Corollaries and Extensions}

\begin{corollary}
For \( \delta = 0.1 \), \( d_0 \approx 1.8 \log_{10} t_\text{max} \); for \( h = 128 \), \( q_\text{max} \leq 2^{127} \). Practical anti-admissibility is achievable.
\end{corollary}

\begin{corollary}
Extending to \( n \)-out-of-\( m \) threshold keys increases \( h_0 \) by \( \log \binom{m}{n} \), enabling distributed guardianship.
\end{corollary}

\section{Conclusion}

The theorem rigorously proves that composed ritual-cryptographic resistances suffice for anti-admissibility, offering a formal escape from Ellul's technological closure within resource-bounded regimes.

\end{document}
