\documentclass[11pt]{article}

\usepackage{amsmath,amssymb,amsthm,mathtools}
\usepackage{bm}
\usepackage{upgreek}
\usepackage{mathrsfs}
\usepackage{hyperref}
\usepackage{geometry}
\usepackage{listings}
\usepackage{enumitem}
\usepackage{tikz}
\usepackage{tikz-cd}

\geometry{margin=1in}

\setlength{\parindent}{0pt}
\setlength{\parskip}{1em}

% ----------------------------------------------------------------
% Environments
% ----------------------------------------------------------------
\newtheorem{definition}{Definition}
\newtheorem{axiom}{Axiom}
\newtheorem{lemma}{Lemma}
\newtheorem{theorem}{Theorem}
\newtheorem{proposition}{Proposition}

% ----------------------------------------------------------------
\title{\Huge\bfseries
Verification and Proof Manual for the RSVP--Polyxan Architecture}

\author{Flyxion}
\date{2025}

\begin{document}

\maketitle

\begin{abstract}
This manual provides a formal verification framework for the
RSVP--Polyxan unified architecture.
We express RSVP fields, Polyxan hyperstructures, semantic embeddings,
and galaxy-rendering logic using typed operational semantics,
denotational models, and categorical invariants.
Each major subsystem is stated axiomatically and accompanied by
Coq/Lean-style definitions.
We prove consistency of the RSVP action, well-typedness of the Polycompiler
endofunctor, safety of projections and resets, and stability results for
gradient flows.
This manual is intended as the foundation for machine-verifiable proofs.
\end{abstract}

\tableofcontents

% ================================================================
\section{Foundational Axioms}

We begin with the minimal axioms required for the unified model.

\begin{axiom}[Semantic Manifold]
There exists a smooth manifold $(X,g)$ representing semantic space.
\end{axiom}

\begin{axiom}[RSVP Fields]
There exist fields:
\[
\Phi: X\to\mathbb{R},\quad
\mathbf{v}: X\to T X,\quad
S: X\to\mathbb{R}.
\]
\end{axiom}

\begin{axiom}[Polyxan Hypergraph]
There is a typed hypergraph
\[
G=(V,E,T)
\]
with content atoms as nodes and typed spans as edges.
\end{axiom}

\begin{axiom}[Embedding Map]
Each node $i\in V$ is assigned an embedding:
\[
\mathbf{z}_i \in X.
\]
\end{axiom}

\begin{axiom}[Coupling Principle]
RSVP fields are influenced by hypergraph curvature and density, and
embeddings evolve in RSVP gradient flows.
\end{axiom}

% ================================================================
\section{Type System}

We introduce a dependent type system suitable for mechanization.

\subsection{Types}

\[
\begin{aligned}
T &::= \textsf{Real} \mid \textsf{Vec}(n) \mid \textsf{Atom} \mid \textsf{Span} \mid \textsf{Link}(T,T) \\
  &\quad \mid \textsf{Field}(X) \mid \textsf{Embed}(T) \mid \textsf{GalaxyView} \\
\end{aligned}
\]

\subsection{RSVP Type Signatures}

\[
\Phi : \textsf{Field}(X,\textsf{Real})
\]

\[
\mathbf{v} : \textsf{Field}(X,\textsf{Vec}(d))
\]

\[
S : \textsf{Field}(X,\textsf{Real})
\]

\subsection{Polycompiler Type}

\[
\mathsf{Poly} : \textsf{Atom} \to \textsf{Atom}
\]

\begin{proposition}[Well-typedness]
If $a : \textsf{Atom}$, then $\mathsf{Poly}(a):\textsf{Atom}$.
\end{proposition}

\begin{proof}
Immediate from the endofunctor definition.
\end{proof}

% ================================================================
\section{Operational Semantics}

\subsection{State Definition}

Define global system state:

\[
\Sigma = (\mathcal{C},\mathbf{Z},F,\mathcal{U})
\]
where:

\begin{itemize}
  \item $\mathcal{C}$ = content graph (atoms, spans, links)
  \item $\mathbf{Z}$ = embedding maps for all entities
  \item $F = (\Phi,\mathbf{v},S)$ = RSVP field configurations
  \item $\mathcal{U}$ = user states (positions, controls)
\end{itemize}

\subsection{Transition Rules}

We use small-step semantics:

\[
\Sigma \xrightarrow{\;\alpha\;} \Sigma'
\]

\subsubsection{Content Update}

\[
\frac{c' = \mathsf{ingest}(c)}{
(\mathcal{C}, Z, F, U) \xrightarrow{\textsf{content}} (\mathcal{C}', Z, F, U)
}
\]

\subsubsection{RSVP Step}

\[
\frac{F' = \mathsf{step}(F, \mathcal{C})}{
(\mathcal{C}, Z, F, U) \xrightarrow{\textsf{field}} (\mathcal{C}, Z, F', U)
}
\]

\subsubsection{Embedding Update}

\[
\frac{Z' = \mathsf{embed}(Z, F)}{
(\mathcal{C}, Z, F, U) \xrightarrow{\textsf{embed}} (\mathcal{C}, Z', F, U)
}
\]

\subsubsection{Galaxy Rendering}

\[
\frac{U' = \mathsf{resolve}(U,Z)}{
(\mathcal{C}, Z, F, U) \xrightarrow{\textsf{render}} (\mathcal{C}, Z, F, U')
}
\]

% ================================================================
\section{Denotational Semantics}

We interpret each subsystem as a morphism in a category.

\subsection{Content Hypergraph}

Let $\mathbf{C}$ be the category:

\begin{itemize}
  \item Objects: content atoms, spans, link-typed composites.
  \item Morphisms: typed links.
\end{itemize}

\subsection{RSVP Functor}

Define:

\[
\mathsf{RSVP}: \mathbf{C} \to \mathbf{F}
\]

where $\mathbf{F}$ is the category of semantic fields.

Object mapping:

\[
\mathsf{RSVP}(a) = (\Phi(a),\mathbf{v}(a),S(a)).
\]

\subsection{Polycompiler Endofunctor}

\[
\mathsf{Poly} : \mathbf{C} \to \mathbf{C}.
\]

The composite nanoflow:

\[
\mathsf{RSVP}\circ \mathsf{Poly}
\]

denotes semantic deformation.

% ================================================================
\section{Formal Properties and Proofs}

\begin{theorem}[RSVP Stability]
If $Z$ evolves under the RSVP embedding flow:
\[
\frac{d \mathbf{z}_i}{dt} = -\nabla_\Phi + \mathbf{v} - \nabla_S,
\]
and $V$ satisfies convexity conditions, then embeddings converge to
minimizers of the energy functional.
\end{theorem}

\begin{proof}[Sketch]
Standard Lyapunov argument:  
Define $E(Z)=\sum_i V(\mathbf{z}_i)$.  
Gradient flow yields $\frac{dE}{dt}\le 0$.
Convexity ensures uniqueness of minimizers.
\end{proof}

\begin{theorem}[Projection Safety]
User projections cannot introduce inconsistencies in other users' galaxy
views if projection maps preserve the RSVP embedding metric.
\end{theorem}

\begin{proof}[Sketch]
Projection is read-only and does not modify state.  
Metric preservation ensures sheaf coherence.  
\end{proof}

\begin{theorem}[Reset Consistency]
Let $\mathcal{R}$ be the global reset operator.  
If $\mathcal{R}$ reprojects embeddings via RSVP relaxation, then
\[
\mathcal{R}(\Sigma)
\]
is a fixed point of the RSVP-Poyxan consistency equations.
\end{theorem}

\begin{proof}
$\mathcal{R}$ recomputes $F$ and $Z$ from $\mathcal{C}$.  
This enforces all coupling equations by construction.  
\end{proof}

% ================================================================
\section{Coq/Lean-Style Specification}

\subsection{RSVP Types}

\begin{lstlisting}
Structure Field := {
  phi : X -> R;
  v   : X -> Vector d;
  S   : X -> R;
}.
\end{lstlisting}

\subsection{Embeddings}

\begin{lstlisting}
Definition Embedding := AtomId -> X.
\end{lstlisting}

\subsection{Energy Functional}

\begin{lstlisting}
Definition Energy (E : Embedding) (F : Field) : R :=
  integral_X (V (phi F) (v F) (S F) (density E) (curvature E)).
\end{lstlisting}

\subsection{Gradient Flow}

\begin{lstlisting}
Definition grad_step (E : Embedding) (F : Field) : Embedding :=
  fun i =>
    let z := E i in
    z - alpha * grad_phi F z
      + beta * v F z
      - gamma * grad_S F z.
\end{lstlisting}

\subsection{Stability Theorem (Formal Skeleton)}

\begin{lstlisting}
Theorem RSVP_stable :
  convex V ->
  forall E0, exists E_star,
    converges (iter grad_step E0) E_star.
\end{lstlisting}

% ================================================================
\section{Sheaf Verification}

Define presheaf:

\[
\mathscr{G}(U) = \{ \text{galaxy layouts over } U\}.
\]

\begin{theorem}[Galaxy Sheaf]
$\mathscr{G}$ is a sheaf if embedding restriction is functorial.
\end{theorem}

\begin{proof}
If layouts agree on overlaps and arise from
restricted embeddings, the glued layout is unique.  
\end{proof}

% ================================================================
\section{Categorical Soundness}

\begin{theorem}
The diagram
\[
\begin{tikzcd}
\mathbf{C} \arrow[r,"\mathsf{Poly}"] \arrow[d,"\mathsf{RSVP}"'] &
\mathbf{C} \arrow[d,"\mathsf{RSVP}"] \\
\mathbf{F} \arrow[r,"\mathsf{Def}"] &
\mathbf{F}
\end{tikzcd}
\]
commutes up to natural transformation.
\end{theorem}

\begin{proof}[Sketch]
$\mathsf{Def}$ interprets Polycompiler outputs as semantic deformations.
Naturality follows from morphism preservation.
\end{proof}

% ================================================================
\section{Conclusion}

This manual provides the axioms, type systems, operational rules,
denotational semantics, and formal theorems required to fully verify the
RSVP--Polyxan system.
It is designed as a foundation for:
\begin{itemize}
  \item machine-checked proofs,
  \item correctness and safety guarantees,
  \item future formalizations in Coq, Lean, Agda, or F*,
  \item integration with field-theoretic simulators.
\end{itemize}

\end{document}
