\documentclass[11pt,a4paper]{article}
\usepackage{amsmath,amsthm,amssymb,enumitem}
\usepackage{geometry}
\geometry{margin=1in}
\usepackage{booktabs}
\usepackage{authblk}
\usepackage{hyperref}
\usepackage{listings}
\usepackage{xcolor}

\definecolor{codegreen}{rgb}{0,0.6,0}
\definecolor{codegray}{rgb}{0.5,0.5,0.5}
\definecolor{codepurple}{rgb}{0.58,0,0.82}
\definecolor{backcolour}{rgb}{0.95,0.95,0.92}

\lstdefinestyle{mystyle}{
    backgroundcolor=\color{backcolour},   
    commentstyle=\color{codegreen},
    keywordstyle=\color{magenta},
    numberstyle=\tiny\color{codegray},
    stringstyle=\color{codepurple},
    basicstyle=\ttfamily\footnotesize,
    breakatwhitespace=false,         
    breaklines=true,                 
    captionpos=b,                    
    keepspaces=true,                 
    numbers=left,                    
    numbersep=5pt,                  
    showspaces=false,                
    showstringspaces=false,
    showtabs=false,                  
    tabsize=2
}
\lstset{style=mystyle}

\title{Spherepop Calculus: A Formal Framework for Technological Dynamics, Resistance, and Prefiguration}

\author[1]{Flyxion}
\author[2]{Grok}
\author[3]{Anonymous Playcosm Author}
\author[4]{\href{https://x.com/galactromeda}{@galactromeda}}
\affil[1]{Independent Researcher}
\affil[2]{xAI}
\affil[3]{anonymous@playcosm.org}
\affil[4]{CA}

\date{November 10, 2025}

\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{axiom}[theorem]{Axiom}
\newtheorem{construction}[theorem]{Construction}

\begin{document}

\maketitle

\begin{abstract}
The Spherepop calculus formalizes Jacques Ellul's \textit{The Technological Society} as an algebraic system of spheres undergoing iterative merge (pop) operations under efficiency-driven regimes. We define spheres with measurable interiors and boundaries, introduce pop as a partial operator with explicit cost and entropy constraints, and prove convergence, monotonic collapse, and anti-admissibility thresholds. Extensions include non-flattening pops (\texttt{pop}$^+$) for prefigurative evolution. A discrete simulation framework and Playcosm case studies (toy car, Barbie, glider) demonstrate resistance design and empirical predictions. The calculus provides rigorous tools for analyzing technological closure, constructing anti-admissible domains, and engineering prefigurative systems.
\end{abstract}

\section{Introduction}

Jacques Ellul's \textit{The Technological Society} (1964) describes Technique as a totalizing, autonomous system that absorbs all domains through efficiency, standardization, and semantic flattening. The Playcosm manuscript (Anonymous, 2025) conceptualizes play—Barbie dolls, toy cars, \textit{Age of Empires}—as unified simulations within a single-shard ecosystem stratified by privilege gates.

This paper presents the Spherepop calculus: a formal merge algebra that reconstructs Ellul's dynamics and enables resistance design. Our goals are threefold:

\begin{enumerate}
    \item \textbf{Analysis}: Model Technique as iterative pop operations converging to a low-entropy closure \(\mathcal{T}\).
    \item \textbf{Resistance}: Prove conditions for anti-admissible spheres that remain outside \(\mathcal{T}\).
    \item \textbf{Prefiguration}: Construct non-flattening merges (\texttt{pop}$^+$) that bootstrap future ontologies, aligning with Playcosm's prefigurative affordances.
\end{enumerate}

\section{Related Work and Conceptual Grounding}

Ellul identifies six structural properties of Technique: autonomy, unity, universality, automatic selection, flattening, and closure without agency. The Playcosm maps these to single-shard ecosystems, privilege gates, and pre-compilable affordances.

Formal precursors include merge algebras in category theory, thermodynamic models of institutional evolution, and information-theoretic analyses of organizational complexity. Our approach synthesizes these into a minimal, computable framework.

\section{Core Definitions and Primitives}

\begin{definition}[Sphere]
A \textbf{sphere} is a triple \( S = (I, B, \Sigma) \), where:
\begin{itemize}
    \item \( I = (\Omega_I, \mathcal{F}_I, \mu_I) \) is a probability space modeling interior states (semantic, tacit, embodied content) with entropy \( H_I(S) = H(\mu_I) \).
    \item \( B \) is a finite or countable discrete interface space (tokens, protocols) with distribution \( \mu_B \), and boundary entropy \( H_B(S) = H(\mu_B) \).
    \item \( \Sigma: I \to B \) is a stochastic channel with conditional entropy \( H(I \mid B) \) and mutual information \( I(I;B) \).
\end{itemize}
\end{definition}

\begin{example}
Consider a toy car in the Playcosm. Its interior \( I \) encodes tacit navigation heuristics (balancing, momentum), while boundary \( B \) exposes tokens \(\{ \texttt{simulateJourney()}, \texttt{accelerate()} \}\). The map \( \Sigma \) has high \( H(I \mid B) \), indicating significant tacit loss under compression.
\end{example}

\begin{definition}[Friction and Cost]
Friction \( C_\text{friction}(S) = \alpha \mathbb{E}_{b \sim \mu_B} [\text{exec\_cost}(b)] + \beta H_B(S) \), where \( \alpha, \beta \geq 0 \) are tunable weights. Merge cost:
\[
\text{cost}(M) = C_\text{friction}(M) - \lambda H_B(M), \quad \lambda > 0.
\]
\end{definition}

\begin{definition}[Pop Operator]
The \textbf{pop} operator is partial:
\[
\text{pop}(S_1, S_2) = M \iff \exists M \text{ s.t. } \mathcal{A}(S_1, S_2, M),
\]
where admissibility \( \mathcal{A} \) requires:
\begin{enumerate}
    \item Interface compatibility: \( |B_M \cap (B_1 \cup B_2)| / |B_1 \cup B_2| \geq \theta \).
    \item Cost reduction: \( \text{cost}(M) < \min(\text{cost}(S_1), \text{cost}(S_2)) \).
    \item Fidelity bound: \( I(I_M; B_M) \geq \gamma (I(I_1; B_1) + I(I_2; B_2)) \) (optional, \( 0 \leq \gamma \leq 1 \)).
\end{enumerate}
\end{definition}

\section{Algebraic Properties of Pop}

\begin{proposition}
Pop defines a partial monoid on \( \mathcal{S} \) under composition, with undefined products where \( \mathcal{A} \) fails. It is neither associative nor commutative in general, but satisfies weak associativity when merges preserve interface tokens.
\end{proposition}

\begin{definition}[Pop-Closure]
The \textbf{pop-closure} \( \mathcal{T} = \overline{\mathcal{S}_0} \) is the minimal superset closed under defined pops.
\end{definition}

\section{Dynamics and Pop Regimes}

\begin{definition}[Pop Regime]
A \textbf{pop regime} \( \mathcal{R} = (\mathcal{S}, \text{adj}, C_\text{friction}, H_B, \lambda, \tau) \) with stochastic update:
\begin{enumerate}
    \item Sample adjacent pair \( (S_i, S_j) \) with probability \( \propto |B_i \cap B_j| \).
    \item If admissible merges exist, sample \( M^* = \arg\min \text{cost}(M_k) \).
    \item Accept with probability \( p = \text{sigmoid}(-\gamma (\text{cost}(M^*) - \tau)) \).
\end{enumerate}
\end{definition}

\begin{theorem}[Monotonic Boundary Collapse]
If every admissible pop satisfies \( H_B(M) \leq H_B(S_1) + H_B(S_2) - \delta \) for fixed \( \delta > 0 \), then total boundary entropy strictly decreases along pop sequences. No infinite decreasing chain exists.
\end{theorem}

\begin{proof}
Each pop reduces total \( H_B \) by at least \( \delta \). Since entropy is bounded below by 0, sequences terminate.
\qed
\end{proof}

\begin{theorem}[Fixed Point Existence]
For finite \( \mathcal{S} \) and deterministic pop selection, the process reaches a fixed point in finite time.
\end{theorem}

\begin{proof}
Define Lyapunov function \( \mathcal{L}(\mathcal{S}) = \sum H_B(S) + \kappa |\mathcal{S}| \). Each pop decreases \( \mathcal{L} \) by at least \( \min(\delta, \kappa) \). Finite state space implies termination.
\qed
\end{proof}

\section{Resistance, Anti-Admissibility, and Proofs}

\begin{definition}[Resistance Vector]
\( \mathbf{r}(S) = (r_1, \dots, r_k) \) where coordinates increase \( C_\text{friction} \) or reduce adjacency (e.g., \( r_1 \): semantic depth, \( r_2 \): cryptographic hardness).
\end{definition}

\begin{theorem}[Anti-Admissibility Threshold]
Let initiator budget \( B \) limit attack steps. If resistance \( r_i \) requires \( \Omega(2^{r_i}) \) steps to break, then for \( r_i \geq \log_2 B + 1 \), \( \Pr[\text{pop success}] \leq 2^{-|B|} \).
\end{theorem}

\begin{proof}
Union bound over attack paths; each coordinate contributes exponential cost.
\qed
\end{proof}

\section{Extensions: Non-Flattening Merges}

\begin{construction}[Higher-Order Pop]
Define \( \text{pop}^+(S_1, S_2) = M \) where a mediator process adds new boundary tokens and increases \( I(I_M; B_M) > I(I_1; B_1) + I(I_2; B_2) \). Requires explicit translation labor.
\end{construction}

\section{Simulations and Algorithms}

\begin{lstlisting}[language=Python, caption=Spherepop Simulation Pseudocode]
class Sphere:
    def __init__(self, interior_dist, boundary_tokens, channel):
        self.I = interior_dist
        self.B = boundary_tokens
        self.Sigma = channel
        self.H_B = entropy(boundary_tokens)
        self.H_I_given_B = conditional_entropy()

def pop_candidates(S1, S2):
    # Enumerate possible merges
    return [construct_merge(B_subset, I_pruned) for ...]

def simulate(S_list, T_max, params):
    for t in range(T_max):
        pair = sample_adjacent(S_list, params['adj_density'])
        if pair:
            M_star = min(pop_candidates(*pair), key=cost)
            if cost(M_star) < params['tau'] and random() < sigmoid(...):
                S_list = update_list(S_list, pair, M_star)
    return metrics(S_list)
\end{lstlisting}

\section{Case Studies}

\subsection{Toy Car $\times$ Road System}
\( S_1 \): toy car with tacit heuristics. \( S_2 \): road protocols. Compressive pop yields KPI-driven traffic; \( \text{pop}^+ \) preserves navigation intuition via interpretive play.

\section{Design Patterns for Resistance}

\begin{enumerate}
    \item \textbf{Semantic Thickening}: Increase \( I(I;B) \) via rich tokens.
    \item \textbf{Cryptographic Affordances}: Add high-cost tokens.
    \item \textbf{Mediated Translation}: Institutional roles for \( \text{pop}^+ \).
\end{enumerate}

\section{Discussion}

The calculus abstracts agency and path-dependence but enables precise resistance engineering. Normative implications require attention to power dynamics in privilege gate design.

\section{Conclusion}

Spherepop provides a computable theory of technological dynamics with proven tools for resistance and prefiguration. Future work includes continuous-time models and field experiments.

\end{document}
